---
title: "Inference 2 - Hypothesis Testing"
output:
  ioslides_presentation:
    widescreen: yes
  slidy_presentation: default
  beamer_presentation:
    theme: metropolis
---

```{r setup, include=FALSE }
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(fig.height=4)
knitr::opts_chunk$set(fig.width=10)
```

## Introduction

This section covers a fundamental part of inference: hypothesis testing. 

Tests of hypotheses are frequently applied in econometrics, e.g. t-tests for OLS parameters or in tests for heteroscedasticity.

In this section we will:

- Take a look at the general hypothesis testing procedure
- Derive hypothesis testing with a simulation
- Learn how to do hypothesis testing in R

## Introduction

- In the previous part I assumed that we know the population. Actually -- in practice -- we don't

- We use hypothesis testing because we don't much about our populations.

**We actually don't know**:

- The population mean $\mu_X$ (our parameter of interest)
- The population variance $\sigma^2_X$ and standard deviation $\sigma_X$
- The true distribution of the population $X \sim ?(?)$

## Introduction

In this scenario (which is the realistic scenario) we do the following:

1. **Assume the population distribution and a value for our parameter of interest**
2. **Then estimate the unknown parameters from a sample**
3. **Test if our assumptions are reasonable**

 

## Formal Hypothesis Testing Procedure

Assume that the distribution of the population is normal (e.g. because it is reasonably from theory)

1. Formulate a hypothesis about our population parameter (e.g. $H_0:\mu_X = 1000$)
    1.1 Also the alternative hypothesis $H_1: \mu_X \neq 1000$
2. Estimate the parameter
    - In this case with the arithmetic mean $\bar{X}$
3. Estimate the variance and derive the sample standard deviation $S_X$
4. Standardise $\bar{X}$ using our estimated $S_X$ and the assumed $\mu_X$
5. Calculate the probability that our observed $\bar{X}$ is from the population, given a specific level of certainty (e.g. $95 \%$)

## Simulated Population

We take again our sample from a simulated population of ZU student income:

```{r echo=TRUE, message=FALSE, warning=FALSE}
library(ggplot2)
library(tidyverse)
set.seed(11) # seed for reproducibility

n <- 1200
inc_2 <- rnorm(n, mean = 1000, sd = 200)
```

## Simulated Population

```{r}
ggplot() +
geom_histogram(aes(x = inc_2,
y = ..density..),binwidth = 60,alpha = 0.8) +
geom_density(aes(x = inc_2,
y = ..density..),col = "red",size = 2,alpha = 0.8) +
labs(title = "Income of ZU Students") +
theme_minimal()

```

## Sample

We draw a random sample with size `50` from our population:

```{r echo=TRUE}
set.seed(24) ;   sample_n <- 50
sample_2 <- sample(x = inc_2, size = sample_n, replace = F)
```


```{r}
ggplot() +
geom_histogram(aes(x = sample_2,
y = ..density..),binwidth = 50,alpha = 0.8) +
labs(title = "SAMPLE Histogram Income ZU Students") +
theme_minimal()
```



## 1. Formulate the Null-Hypothesis that mean student income is less then $940 €$:
$$ H_0 : \mu_{inc} < 940  $$
Which leaves us with the alternative Hypothesis that the mean income is more or equal to $940€$:
$$ H_1: \mu_{inc} \geq 940 $$
```{r echo=TRUE}
mu_inc = 940
mu_inc
```

## 2. Estimate the mean with the sample mean $\bar{X}$ :
```{r echo=TRUE}
inc_bar <- mean(sample_2)
inc_bar
```

##3. Estimate the sampling variance $S^2_{inc}$ and derive the sampling standard deviation $S_{inc}$ :
```{r echo=TRUE}
var(sample_2)
```
```{r echo=TRUE}
S_inc <- sd(sample_2)
S_inc
```

## Estimated Distribution

At this state we would expect our sample mean to be distributed like this:

```{r}
ggplot(data =  data.frame(X_bar = 850:1030), aes(x=X_bar)) +
  stat_function(fun = dnorm, args = list(mean = mu_inc, sd = S_inc/sqrt(sample_n)), size = 2) +
  geom_vline(xintercept = inc_bar, color = "red", size = 1.5) +
  geom_vline(xintercept = mu_inc, color = "green", size = 1.5) +
  labs(title = "Assumed PDF of our sample mean under H0") +
  theme_minimal()
```

Note that we used the *assumed* mean ZU student income $\mu_{inc} =940€$ (green line). If our $H_0$ would be true, this would be our estimator distribution.

## 4. Standardise $\bar{X}$ to calculate probabilities

```{r echo=TRUE}
Z_inc_bar <- (inc_bar - 940) / (S_inc / sqrt(sample_n)) ; Z_inc_bar
```

```{r}
ggplot(data =  data.frame(Z_inc_bar = -4:4), aes(x=Z_inc_bar)) +
  stat_function(fun = dnorm, args = list(mean = 0, sd = 1), size = 2) +
  geom_vline(xintercept = Z_inc_bar, color = "red", size = 1.5) +
  geom_vline(xintercept = 0, color = "green", size = 1.5) +
  labs(title = "Assumed PDF of our standardised sample mean under H0") +
  theme_minimal()
```

##5. Calculate probability of $\bar{X}$ occuring at 10% significance level

Look up $z-value$ for $1-10\% = 90\%$, e.g. in this [table](http://www.z-table.com/). $z_{90\%} = 1.65$

```{r}
ggplot(data =  data.frame(Z_inc_bar = -4:4), aes(x=Z_inc_bar)) +
  stat_function(fun = dnorm, args = list(mean = 0, sd = 1), size = 2) +
    stat_function(fun = dnorm, xlim = c(-4,1.65), geom = "area", fill = "grey", alpha=0.5) +
  stat_function(fun = dnorm, xlim = c(1.65,4), geom = "area", fill = "black", alpha=0.5) +
  geom_vline(xintercept = Z_inc_bar, color = "red", size = 1.5) +
  geom_vline(xintercept = 0, color = "green", size = 1.5) +
  labs(title = "Assumed PDF of our standardised sample mean under H0") +
  theme_minimal()
```
