---
title: "Using the tidyverse"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data Manipulation


1. We learned that we can read data in the .csv format using `read.csv` and `read.csv2`, and we also got to know the `haven` library. Load the dataset "ISSP.dat" using one of `haven`'s `read_` functions and store it in a variable.

***

2. The ISSP dataset consists of many single observations from many countries. We are only interested in observations from Germany. The variable **V3** signifies the country, the value 276.1 stands for West Germany, the value 276.2 for East Germany. Using `dplyr`'s `filter()`, select only the rows containing observations from Germany. Store the resulting table in the variable "Germany_raw".

***

3. Now, we want to do some subsequent data manipulation.
    a. First, the dataset still consists of unnecessarily many variables. We only want to look at the variables "sex", "age", and "DE_RINC" (Real Income). Use `dplyr`'s `select()` to select only the variables in question.
    b. Looking at the result, it becomes apparent that the "DE_RINC" variable is stored as a string. Since we want to work with numbers, we need to convert the variable to a numeric format. Use `mutate()` to redefine the "DE_RINC" variable by calling `as.numeric()` on it.
    c. We have also noted that we cannot really infer anything from our sex variable, since it only consists of 1s and 2s. Use `mutate()` and `factor()` to convert sex to a factor with the levels / labels 1="m", 2="w".
    d. Lastly, we want to create a dichotomous variable "highInc" that indicates whether the person in question has a monthly real income of 5000 or more. Use `mutate()` in combination with `if_else` (or if you like: `case_when()`) to create this variable.
    e. Store the resulting table in a variable called "Germany"
  
***

4. Have a look at the share of men and women who have a "high income". Of all men, what is the percentage of men with a high income? And of all women, what is the percentage of women with a high income? Use `group_by()` and `summarise()` to calculate the mean of the "highInc" variable within each group. Finally, arrange the resulting table by your percentage variable (ascending). 

*** 

5. 
    a. Define a new variable "Australia_raw" and filter the ISSP dataset so that it only contains observations from Australia (V3==36). Then use `bind_rows()` to merge it with the existing Germany_raw dataframe. Store this merged dataframe into the variable "AUSGER" (don't worry about the warnings).
    b. You notice that you have forgotten a column for your "Germany" dataframe. Use `bind_cols()` to paste the "degree" column of the "Germany_raw" dataframe onto the "Germany" dataframe and store it in the variable "Germany"!
    c. Use `mutate()` to generate a new "ID" column with a running index (1,2,...,`n()`) for your "Germany" dataframe. Then use the following code to generate a dataframe comprising random data:
```{r echo=T, eval=F}
randomData <- data_frame(ID=1:nrow(Germany), 
                         rand=sample(1:1000, nrow(Germany), replace=T))
```
Now use `left_join` to join your "Germany" and "randomData" dataframes by the ID column!
